Shallow EDSLs and Object-Oriented Programming

### 2016-06-11 paper/paper.tex
abstract

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a Domain Specific
Language into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). Given
that Object-Oriented Languages have evolved over more than 50 years 
to optimize the use of procedural abstraction, they ought to have some 
advantages to encode shallow EDSLs.

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.
To alleviate some of the syntactical disadvantages of Java, we create
an annotation that uses transparent code generation techniques to
automatically eliminate large portions of boilerplate code in Java.
To further illustrate the applicability of our techniques, we conduct 
several case studies using larger DSLs from the literature.

>>>>

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a Domain Specific
Language into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). Given
that Object-Oriented Languages have evolved over more than 50 years 
to improve the use of procedural abstraction, they ought to have some 
advantages to encode shallow EDSLs.

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.
To alleviate some of the syntactical disadvantages of Java, we create
an annotation that uses transparent code generation techniques to
automatically eliminate large portions of boilerplate code.
To further illustrate the applicability of our techniques, we conduct 
several case studies using larger DSLs from the literature.
